Single cell resolution landscape of equine peripheral blood mononuclear cells reveals diverse cell types including T-bet+ B cells.

### 2020-12-11 analysis/paper/paper.Rmd
Commit code updates

# Set up R environment
```{r echo = FALSE, message=FALSE}
### Load packages necessary for script
library(renv)
library(sctransform)
library(Matrix)
library(AnnotationDbi)
library(tibble)
library(gtools)
library(cowplot)
library(gdata)
library(clustree)
library(future)
library(rrtools)
library(rlang)
library(Seurat)
library(ggnetwork)
library(RColorBrewer)
library(ggrepel)
library(readxl)
library(ggpubr)
library(lemon)
library(scales)
library(edgeR)
library(dplyr)
library(patchwork)
library(ape)
library(mgsub)
library(jntools)
library(BiocManager)
library(viridis)
library(ggtree)
library(ggforce)
library(legocolors)
library(openxlsx)
library(cluster)
library(factoextra)
library(GEOquery)
library(ComplexHeatmap)
library(circlize)
library(here)
library(reshape2)

### Load custom functions used in this analysis
source(here("R/custom_functions.R"))
### Point to project directory
project_dir <- here("equinepbmc/")
### Set paths for figures and data outputs
figuresPath <- paste0(project_dir, "/analysis/figures/")
derivedDataPath <- paste0(project_dir, "/analysis/data/derived_data/")
### Read on color-annotation file
colorSchema_ePBMC <- read_excel(path = here("analysis/data/derived_data/colorschema_annotations.xlsx"))
```

@@

```{r Reading in the ESAT processed gene-cell matrices}
## Pull counts matrices directly from GEO
tmpdir <- tempdir()
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE148416&format=file"
file <- basename(url)
download.file(url, file)
untar(file, compressed = "gzip", exdir = tmpdir)

gz.files <- lapply(list.files(tmpdir, full.names = T, pattern = "GSM4468652"), gzfile)
counts <- lapply(gz.files, read.table, sep = "\t")
names(counts) <- strsplit(list.files(data_dir), split = "_") %>% sapply("[[", 2)
```

@@

```{r Apply DoubletFileConverter to doublet files}
doublet_out_dir <- here("analysis/data/raw_data/DoubletDetector_output")
doublet_data <- lapply(list.files(doublet_out_dir, full.names = T), read.table, row.names = 1)
names(doublet_data) <- strsplit(list.files(doublet_out_dir), split = "_") %>% sapply("[[", 1)
doublet_data <- lapply(doublet_data, function(x) {
  rownames(x) <- gsub(x = rownames(x), pattern = ":", replacement = ".")
  colnames(x) <- "doublet"
  return(x)
})
```

@@

```{r Qauntify mitochondrial RNA contributions in the single cells in all 7 subjects}
seurat.obj <- lapply(seurat.obj, function(object) {
  object[["percent.mito"]] <- PercentageFeatureSet(object, features = toMatch)
  return(object)
})
```

@@

```{r Perform filtering of Seurat objects against user-defined thresholds}
seurat.obj <- lapply(seurat.obj,
  subset,
  subset =
    nCount_RNA < 20000 & nCount_RNA > 750 &
      nFeature_RNA > 200 &
      doublet < 0.5 &
      percent.mito < 5
)
```

@@

```{r Independent normalizaiton (SCTransform) for each subject}
## Previous analyses showed clustering/gene expression patterns highly influenced by mitochondrial gene expression, thus we regress it at this step.
seurat.obj <- lapply(seurat.obj, SCTransform, vars.to.regress = "percent.mito", verbose = FALSE)
```

@@

immunoglobulin.genes <- c(
  grep(pattern = "^IGH", x = rownames(seurat.obj$s1), value = T)[-1],
  "ENSECAG00000031522", "ENSECAG00000001923",
  "ENSECAG00000039599", "ENSECAG00000038455",
  "ENSECAG00000031094", "ENSECAG00000015109",
  "ENSECAG00000033354", "ENSECAG00000007258"
)

@@

```{r Dimensional Reduction via Principal component analysis}
ePBMC.integrated.obj <- RunPCA(
  object = ePBMC.integrated.obj,
  npcs = 100, verbose = TRUE,
  features = SelectedFeatures
)
```

@@

## Heatmaps to explore the principal components
## Set number of dims to visualize
ndims <- 100
## Set number of dimensions to visualize per plot
ndims.plot <- 6
for (i in 1:as.integer(ndims / ndims.plot)) {
  dims <- seq_range(x = 1:ndims, by = 5)[i:(i + 1)]
  DimHeatmap(object = ePBMC.integrated.obj, dims = dims[1]:dims[2], cells = 100, fast = T, balanced = T)
}
```

@@

# Run UMAP embedding
ePBMC.integrated.obj <- RunUMAP(ePBMC.integrated.obj, dims = 1:25, min.dist = 0.5)

# Build hierarchical clustering based on clustering results
ePBMC.integrated.obj <- BuildClusterTree(ePBMC.integrated.obj, assay = "RNA", dims = 1:25)
```

@@

```{r Filtering step: Remove putative doublets based on knowledge of immune cell development}
## FeatureScatters
# CD3G MS4A1 double positive cells
CD3G_MS4A1.doublets.plot <- FeatureScatter(ePBMC.integrated.obj,
  feature1 = "CD3G",
  feature2 = "MS4A1"
)
# CD3G CD14 double positive cells
CD3G_CD141.doublets.plot <- FeatureScatter(ePBMC.integrated.obj,
  feature1 = "CD3G",
  feature2 = "CD14"
)
# MS4A1 CD14 double positive cells
MS4A1_CD14.doublets.plot <- FeatureScatter(ePBMC.integrated.obj,
  feature1 = "MS4A1",
  feature2 = "CD14"
)

ePBMC.integrated.obj <- CellCycleScoring(
  object = ePBMC.integrated.obj,
  s.features = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes, set.ident =
    FALSE
)
reg.plot <- DimPlot(ePBMC.integrated.obj, group.by = "Phase")
# Manually split this cluster up into two clusters: cluster 21-G2M; cluster 21-S
ePBMC.integrated.obj <- CellSelector(plot = reg.plot, object = ePBMC.integrated.obj, ident = "31")
ePBMC.integrated.obj <- StashIdent(object = ePBMC.integrated.obj, "clusters")
```

@@

```{r Setting up edgeR objects for differential gene expression testing}
## Order the Seurat objects identities from low to high numerically
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "clusters")
cellOrder <- seq(0, 31)
ePBMC.integrated.obj@meta.data$order <- factor(x = ePBMC.integrated.obj$clusters, levels = cellOrder)
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")
ePBMC.integrated.obj$order <- droplevels(ePBMC.integrated.obj$order)

## Detect genes that are expressed in at least 10% of any given cluster in the Seurat object and write them out into a vector called "marker.names".
GeneDetector(SeuratObj = ePBMC.integrated.obj, threshold = 0.10, listname = "marker.names")

## Filter the raw gene-cell counts matrix(stored in the Seurat object) with the genes identified above
tmp.matrix <- ePBMC.integrated.obj@assays$RNA@counts[marker.names, ]

## Create the Seurat object to recalculate the nFeature_RNA (unique genes detected/cell)
tmp <- CreateSeuratObject(tmp.matrix)
colnames(tmp@meta.data)[3] <- "Recalculated_nGene"
tmp <- AddMetaData(object = tmp, metadata = ePBMC.integrated.obj@meta.data)

## Calculate the gdr from the nGene column in the Seurat metadat
tmp@meta.data$gdr <- as.numeric(scale(tmp@meta.data$Recalculated_nGene))
tmp@meta.data$orig.ident <- NULL

## Create DGE counts object from the edgeR package
counts <- DGEList(counts = tmp@assays$RNA@counts, genes = rownames(tmp))
## Calculate the library sizes of each single cell to be analyzed
counts$samples$lib.PBMSCize <- colSums(counts$counts)
## Calculate the normalization factors to be applied to make single cell comparable
counts <- edgeR::calcNormFactors(counts)

## Reformat metadata from the Seurat object to make compatible with edgeR/design matrix
edgeRmetadata <- tmp@meta.data[, c("horse", "order", "gdr")]
colnames(edgeRmetadata)[2] <- c("celltype")
design <- model.matrix(~ 0 + gdr + celltype + horse, data = edgeRmetadata)

## Estimate dispersion
counts <- estimateDisp(counts, design, robust = T)
## Perform the edgeR fit
fit <- glmQLFit(counts, design)
```

@@

```{r Generate constrasts major groups for equine PBMC analysis}
# CD3+ PRF1- lymphocytes
contrastConstructor(
  clusterIDs = c(2, 7, 8, 31, 16, 1, 13, 10, 21, 3, 6),
  design = design,
  name = "CD3PRF1negVALL.constrast",
  makeContrast = T
)
# CD3+ PRF1+ lymphocytes
contrastConstructor(
  clusterIDs = c(4, 5, 14, 17, 20),
  design = design,
  name = "CD3PRF1posVALL.constrast",
  makeContrast = T
)
# B cells
contrastConstructor(
  clusterIDs = c(0, 9, 15, 22, 25, 30, 27),
  design = design,
  name = "BcellVAll.contrast",
  makeContrast = T
)
# Myeloid cells
contrastConstructor(
  clusterIDs = c(11, 12, 18, 28, 19, 24, 26, 23, 29),
  design = design,
  name = "McellVAll.contrast",
  makeContrast = T
)
# Monos cells
contrastConstructor(
  clusterIDs = c(11, 12, 18, 28),
  design = design,
  name = "MocellVAll.contrast",
  makeContrast = T
)
# Basophils
contrastConstructor(
  clusterIDs = c(23),
  design = design,
  name = "BasoVAll.contrast",
  makeContrast = T
)
# Dendritic cells
contrastConstructor(
  clusterIDs = c(19, 24, 26),
  design = design,
  name = "DCVAll.contrast",
  makeContrast = T
)

major.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "VAll", value = T), get))
colnames(major.contrasts) <- grep(ls(), pattern = "VAll", value = T)
```

```{r Generate constrasts for subsets within major groups for equine PBMC analysis}
Bcells.cluster.numbers <- c(0, 9, 15, 22, 27, 30)
Mcells.cluster.numbers <- c(11, 12, 18, 28, 19, 24, 26, 29)
Mocells.cluster.numbers <- c(11, 12, 18, 28)
DCcells.cluster.numbers <- c(19, 24, 26)

## B cell subsets
{
  for (i in 1:length(Bcells.cluster.numbers)) {
    contrastConstructor(
      clusterIDs = Bcells.cluster.numbers[i],
      clusterIDs.2 = setdiff(Bcells.cluster.numbers, Bcells.cluster.numbers[i]),
      design = design,
      makeContrast = T,
      name = paste("B", Bcells.cluster.numbers[i], sep = "")
    )
  }

  grep(ls(), pattern = "B[0-9]", value = T)
  B.celltype.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "B[0-9]", value = T), get))
  colnames(B.celltype.contrasts) <- grep(ls(), pattern = "B[0-9]", value = T)
}

## Myeloid cell subsets
{
  for (i in 1:length(Mcells.cluster.numbers)) {
    contrastConstructor(
      clusterIDs = Mcells.cluster.numbers[i],
      clusterIDs.2 = setdiff(Mcells.cluster.numbers, Mcells.cluster.numbers[i]),
      design = design,
      makeContrast = T,
      name = paste("M", Mcells.cluster.numbers[i], sep = "")
    )
  }

  grep(ls(), pattern = "M[0-9]", value = T)
  M.celltype.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "M[0-9]", value = T), get))
  colnames(M.celltype.contrasts) <- grep(ls(), pattern = "M[0-9]", value = T)
}

## Mono cell subsets
{
  for (i in 1:length(Mocells.cluster.numbers)) {
    contrastConstructor(
      clusterIDs = Mocells.cluster.numbers[i],
      clusterIDs.2 = setdiff(Mocells.cluster.numbers, Mocells.cluster.numbers[i]),
      design = design,
      makeContrast = T,
      name = paste("Mo", Mocells.cluster.numbers[i], sep = "")
    )
  }

  grep(ls(), pattern = "Mo[0-9]", value = T)
  Mo.celltype.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "Mo[0-9]", value = T), get))
  colnames(Mo.celltype.contrasts) <- grep(ls(), pattern = "Mo[0-9]", value = T)
}

## DC cell subsets
{
  for (i in 1:length(DCcells.cluster.numbers)) {
    contrastConstructor(
      clusterIDs = DCcells.cluster.numbers[i],
      clusterIDs.2 = setdiff(DCcells.cluster.numbers, DCcells.cluster.numbers[i]),
      design = design,
      makeContrast = T,
      name = paste("DC", DCcells.cluster.numbers[i], sep = "")
    )
  }

  grep(ls(), pattern = "DC[0-9]", value = T)
  DC.celltype.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "DC[0-9]", value = T), get))
  colnames(DC.celltype.contrasts) <- grep(ls(), pattern = "DC[0-9]", value = T)
}
```

@@

```{r Conduct differential gene expression testing for major cell groups and subsets}
major.glmQLFTest <- list()
for (i in 1:length(colnames(major.contrasts))) {
  major.glmQLFTest[[i]] <- glmQLFTest(fit, contrast = major.contrasts[, i]) %>%
    topTags(n = NULL)
}

Myeloid.glmQLFTest <- list()
for (i in 1:length(colnames(M.celltype.contrasts))) {
  Myeloid.glmQLFTest[[i]] <- glmQLFTest(fit, contrast = M.celltype.contrasts[, i]) %>%
    topTags(n = NULL)
}

Mo.glmQLFTest <- list()
for (i in 1:length(colnames(Mo.celltype.contrasts))) {
  Mo.glmQLFTest[[i]] <- glmQLFTest(fit, contrast = Mo.celltype.contrasts[, i]) %>%
    topTags(n = NULL)
}

DC.glmQLFTest <- list()
for (i in 1:length(colnames(DC.celltype.contrasts))) {
  DC.glmQLFTest[[i]] <- glmQLFTest(fit, contrast = DC.celltype.contrasts[, i]) %>%
    topTags(n = NULL)
}

for (i in 1:length(colnames(B.celltype.contrasts))) {
  B.glmQLFTest[[i]] <- glmQLFTest(fit, contrast = B.celltype.contrasts[, i]) %>%
    topTags(n = NULL)
}
```

@@

```{r Heatmap of DEGs between major cell groups}
major_tables <- major.glmQLFTest

major_tables <- major_tables[c("CD3PRF1neg", "CD3PRF1pos", "B", "Monocyte&DC", "Basophil")]
major_tables <-
  lapply(major_tables, tibble::remove_rownames) %>% lapply(tibble::column_to_rownames, "genes")
major.genes <- lapply(
  major_tables,
  function(df) {
    df[order(df$logFC, decreasing = T), ]
  }
) %>%
  lapply(rownames) %>%
  unlist2() %>%
  unique()

@@

```{r Stacked bars}
## Remove contaminating neutrophils
Mcells.sub <- subset(Mcells, idents = 29, invert = T)
Mcells.sub$order <- droplevels(Mcells.sub$order)
horse_mono.dc.cluster_freq.table <- table(Mcells.sub$Subject, Mcells.sub$order) %>% melt()
colnames(horse_mono.dc.cluster_freq.table) <- c("Subject", "Celltype", "Fraction_of_all_cells")
horse_mono.dc.cluster_freq.table$Celltype <- as.factor(horse_mono.dc.cluster_freq.table$Celltype)

@@

```{r  Cross-species dendrogram between human and equine cell type clusters identified by scRNAseq}
# Set identity class to high-resolution clustering results
# Human
hs.integrated.data <- SetIdent(hs.integrated.data, value = "integrated_snn_res.1.2")

# Horse
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")

## Set objects to compare here (seuratObject)
obj_1 <- subset(hs.integrated.data, idents = c(14, 21, 15, 0, 1, 8, 22, 26, 19))
obj_2 <- subset(ePBMC.integrated.obj, idents = Mcells.order[-length(Mcells.order)])
### Extrac there metadata for cluster ids (seuratObject@meta.data)
obj_1.meta <- hs.integrated.data@meta.data
obj_2.meta <- ePBMC.integrated.obj@meta.data
## Set resolution to perform comparisons (STRING)
obj_1.res <- "human.celltype"
obj_2.res <- "order"
### Set string for species parameter for id on dendogram (STRING)
obj_1.species <- "human"
obj_2.species <- "horse"

# Extact counts matrix
obj_1.cMat <- obj_1@assays$RNA@counts
obj_2.cMat <- obj_2@assays$RNA@counts

obj1.Sobj <- CreateSeuratObject(obj_1.cMat)
obj2.Sobj <- CreateSeuratObject(obj_2.cMat)

# Find highly variable genes under the SCTransform framework
obj1.Sobj <- SCTransform(obj1.Sobj)
obj2.Sobj <- SCTransform(obj2.Sobj)

# Extact highly variable genes
obj_1.genes <- VariableFeatures(obj1.Sobj)
obj_2.genes <- VariableFeatures(obj2.Sobj)

# Intersect species hvg's to identify shared highly variable orthologs
orthologs <- intersect(obj_1.genes, obj_2.genes)

# Use RelativeCounts function within Seurat to calculate TPM from counts matrix
obj_1.TPM <- RelativeCounts(obj_1.cMat, scale.factor = 10^6)
obj_2.TPM <- RelativeCounts(obj_2.cMat, scale.factor = 10^6)

# Add some cluster meta data to cell names in the TPM matrices; will be used to identify whose 'who'
obj_1 <- SetIdent(obj_1, value = obj_1.res)
obj_2 <- SetIdent(obj_2, value = obj_2.res)

colnames(obj_1.TPM) <- paste(colnames(obj_1.TPM), Idents(obj_1), sep = "_")
colnames(obj_2.TPM) <- paste(colnames(obj_2.TPM), Idents(obj_2), sep = "_")

# Filter matrices by highly var orthologs to make file sizes more maneagable
obj_1.TPM <- obj_1.TPM[orthologs, ]
obj_2.TPM <- obj_2.TPM[orthologs, ]

# Convert from sparseMatrix to matrix format to work well with base R functions
obj_1.TPM <- as.matrix(obj_1.TPM)
obj_2.TPM <- as.matrix(obj_2.TPM)

# Tidy data and calculate gene-cluster averages across all groups
library(reshape2)
obj_1.melt <- melt(obj_1.TPM)
colnames(obj_1.melt) <- c("gene", "cell", "count")
obj_1.melt$cluster <- sub(".*[0-9]_", "", obj_1.melt$cell)

obj_2.melt <- melt(obj_2.TPM)
colnames(obj_2.melt) <- c("gene", "cell", "count")
obj_2.melt$cluster <- sub(".*_", "", obj_2.melt$cell)

library(dplyr)
obj_1.groupedbyClusterandGene_means <- obj_1.melt %>%
  group_by(cluster, gene) %>%
  summarise(mean = mean(count))
obj_2.groupedbyClusterandGene_means <- obj_2.melt %>%
  group_by(cluster, gene) %>%
  summarise(mean = mean(count))

obj_1.groupedbyClusterandGene_means.df <- dcast(data = obj_1.groupedbyClusterandGene_means, formula = gene ~ cluster, fun.aggregate = sum, value.var = "mean")
rownames(obj_1.groupedbyClusterandGene_means.df) <- obj_1.groupedbyClusterandGene_means.df[, 1]
obj_1.groupedbyClusterandGene_means.df[, 1] <- NULL

obj_2.groupedbyClusterandGene_means.df <- dcast(data = obj_2.groupedbyClusterandGene_means, formula = gene ~ cluster, fun.aggregate = sum, value.var = "mean")
rownames(obj_2.groupedbyClusterandGene_means.df) <- obj_2.groupedbyClusterandGene_means.df[, 1]
obj_2.groupedbyClusterandGene_means.df[, 1] <- NULL

# Add psuedocount TPM of 50 to each matrix of mean expression
obj_1.groupedbyClusterandGene_means.df.p <- obj_1.groupedbyClusterandGene_means.df + 50
obj_2.groupedbyClusterandGene_means.df.p <- obj_2.groupedbyClusterandGene_means.df + 50

# Calculate medians of average expression per gene
obj_1.median.mean <- apply(obj_1.groupedbyClusterandGene_means.df.p, MARGIN = 1, median)
obj_2.median.mean <- apply(obj_2.groupedbyClusterandGene_means.df.p, MARGIN = 1, median)

# Add psuedocount TPM of 50 to each median
obj_1.median.mean.p <- obj_1.median.mean + 50
obj_2.median.mean.p <- obj_2.median.mean + 50

# Perform klein transformation
obj_1.kleinTrans <- sweep(obj_1.groupedbyClusterandGene_means.df.p, 1, obj_1.median.mean, "/")
obj_2.kleinTrans <- sweep(obj_2.groupedbyClusterandGene_means.df.p, 1, obj_2.median.mean, "/")

## Append species identifier to column names
colnames(obj_1.kleinTrans) <- paste(obj_1.species, colnames(obj_1.kleinTrans), sep = "_")
colnames(obj_2.kleinTrans) <- paste(obj_2.species, colnames(obj_2.kleinTrans), sep = "_")

# Combine matrices via bind columns(species clusters)
merged.obj <- cbind(obj_1.kleinTrans, obj_2.kleinTrans)

# Perform log transformation(natural)
merged.obj.log1p <- log1p(merged.obj)

# Calculate pearson distance metrices
pearson.dist.df <- as.dist((1 - cor(merged.obj.log1p)) / 2)

# Run hclust and visualize dendogram
dendo <- hclust(pearson.dist.df, method = "ward.D2")

@@

```{r Cross-species dendrogram between human and equine B cell type clusters identified by scRNAseq}
# Human
hs.integrated.data <- SetIdent(hs.integrated.data, value = "integrated_snn_res.1.2")

# Horse
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")

>>>>

# Set up R environment
```{r echo = FALSE, message=FALSE}
### Load packages necessary for script
library(renv)
library(Matrix)
library(AnnotationDbi)
library(tibble)
library(gtools)
library(cowplot)
library(gdata)
library(clustree)
library(future)
library(rrtools)
library(rlang)
library(uwot)
library(Seurat)
library(sctransform)
library(ggnetwork)
library(RColorBrewer)
library(ggrepel)
library(readxl)
library(ggpubr)
library(lemon)
library(scales)
library(edgeR)
library(dplyr)
library(patchwork)
library(ape)
library(mgsub)
library(jntools)
library(BiocManager)
library(viridis)
library(ggtree)
library(ggforce)
library(legocolors)
library(openxlsx)
library(cluster)
library(factoextra)
library(GEOquery)
library(ComplexHeatmap)
library(circlize)
library(here)
library(reshape2)
library(tidyr)
library(modelr)

### Load custom functions used in this analysis
source(here("R/custom_functions.R"))
### Point to project directory
project_dir <- here("equinepbmc/")
### Set paths for figures and data outputs
figuresPath <- paste0(project_dir, "/analysis/figures/")
derivedDataPath <- paste0(project_dir, "/analysis/data/derived_data/")
### Read on color-annotation file
colorSchema_ePBMC <- as.data.frame(read_excel(path = here("analysis/data/derived_data/colorschema_annotations.xlsx")))
```

@@

```{r Reading in the ESAT processed gene-cell matrices, include=FALSE}
## Pull counts matrices directly from GEO
#tmpdir <- tempdir()
#url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE148416&format=file"
#file <- basename(url)
#download.file(url, file)
#untar(file, compressed = "gzip", exdir = tmpdir)
## Extract and unzip gene-cell matrices only
tmpdir <- "../data/raw_data/GSE148416_RAW/"
gz.files <- lapply(list.files(path = tmpdir, full.names = T, pattern = "GSM"), gzfile)
counts <- lapply(gz.files, read.table, sep = "\t")
names(counts) <- strsplit(list.files(tmpdir), split = "_") %>% sapply("[[", 2)
counts <- lapply(counts, function(x) {
  colnames(x) <- gsub(x = colnames(x), pattern = ".*\\.", replacement = "")
  return(x)
})
```

@@

```{r Apply DoubletFileConverter to doublet files}
doublet_out_dir <- here("analysis/data/raw_data/DoubletDetector_output")
doublet_data <- lapply(list.files(doublet_out_dir, full.names = T), read.table, row.names = 1)
names(doublet_data) <- strsplit(list.files(doublet_out_dir), split = "_") %>% sapply("[[", 1)
doublet_data <- lapply(doublet_data, function(x) {
  rownames(x) <- gsub(x = rownames(x), pattern = ".*:", replacement = "")
  colnames(x) <- "doublet"
  return(x)
})
```

@@

```{r Qauntify mitochondrial RNA contributions in the single cells in all 7 subjects}
seurat.obj <- lapply(seurat.obj, function(object) {
  object[["percent.mito"]] <- PercentageFeatureSet(object, features = toMatch)/100
  return(object)
})
```

@@

```{r Perform filtering of Seurat objects against user-defined thresholds}
seurat.obj <- lapply(seurat.obj,
  subset,
  subset =
    nCount_RNA < 20000 & nCount_RNA > 750 &
      nFeature_RNA > 200 &
      doublet < 0.5 &
      percent.mito < 0.05
)
```

@@

```{r Independent normalizaiton (SCTransform) for each subject}
## Previous analyses showed clustering/gene expression patterns highly influenced by mitochondrial gene expression, thus we regress it at this step.
seurat.obj <- lapply(seurat.obj, 
                     SCTransform, 
                     vars.to.regress = "percent.mito", 
                     verbose = TRUE)
```

@@

immunoglobulin.genes <- c(
  grep(pattern = "^IGH", x = rownames(seurat.obj$s1), value = T)[-1],
    "ENSECAG00000031522", "ENSECAG00000001923",
    "ENSECAG00000039599", "ENSECAG00000038455",
    "ENSECAG00000031094", "ENSECAG00000015109",
    "ENSECAG00000033354"
)

@@

```{r Dimensional Reduction via Principal component analysis}
ePBMC.integrated.obj <- RunPCA(
  object = ePBMC.integrated.obj,
  npcs = 100,
  verbose = TRUE)
```

@@

## Heatmaps to explore the principal components
## Set number of dims to visualize
ndims <- 100
## Set number of dimensions to visualize per plot
ndims.plot <- 6
for (i in 1:as.integer(ndims / ndims.plot)) {
  dims <- modelr::seq_range(x = 1:ndims, by = 5)[i:(i + 1)]
  DimHeatmap(object = ePBMC.integrated.obj, dims = dims[1]:dims[2], cells = 100, fast = T, balanced = T)
}
```

@@

# Run UMAP embedding
ePBMC.integrated.obj <- RunUMAP(ePBMC.integrated.obj,
  dims = 1:25,
  min.dist = 0.6,
  n.neighbors = 75
)

# Build hierarchical clustering based on clustering results
DefaultAssay(ePBMC.integrated.obj) <- "RNA"
ePBMC.integrated.obj <- NormalizeData(ePBMC.integrated.obj)
ePBMC.integrated.obj <- BuildClusterTree(ePBMC.integrated.obj, assay = "RNA", dims = 1:25)
```

@@

```{r Additional step: Identifying cycling cells in the CD3+ lymphocyte compartment}
ePBMC.integrated.obj <- CellCycleScoring(
  object = ePBMC.integrated.obj,
  s.features = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes, set.ident =
    FALSE
)
reg.plot <- DimPlot(ePBMC.integrated.obj, group.by = "Phase")
# Manually split this cluster up into two clusters: cluster 21-G2M; cluster 21-S
ePBMC.integrated.obj <- StashIdent(object = ePBMC.integrated.obj, "clusters")
```

@@

```{r Setting up edgeR objects for differential gene expression testing}
## Order the Seurat objects identities from low to high numerically
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "clusters")
ePBMC.integrated.obj@meta.data$order <- factor(x = ePBMC.integrated.obj$clusters, levels = seq(0, 31))
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")
ePBMC.integrated.obj$order <- droplevels(ePBMC.integrated.obj$order)

## Add Subject meta data
ePBMC.integrated.obj$Subject <- paste0("Subject_", right(rownames(ePBMC.integrated.obj@meta.data), 1))
ePBMC.integrated.obj$Subject <- as.factor(ePBMC.integrated.obj$Subject)

## Add annotation metadata to integrated ePBMC object
ePBMC.integrated.obj$cell_barcode <- rownames(ePBMC.integrated.obj@meta.data)
colorSchema_ePBMC$order <- as.factor(colorSchema_ePBMC$Cluster)
updated_metadata <- right_join(colorSchema_ePBMC, ePBMC.integrated.obj@meta.data, by = "order")
rownames(updated_metadata) <- updated_metadata$cell_barcode
ePBMC.integrated.obj <- AddMetaData(ePBMC.integrated.obj, metadata = updated_metadata)

### Generate edgeR fit, counts DGElist and design matrix
edgeRobj <- edgeR_create_fit(
  object = ePBMC.integrated.obj,
  gene.detection.thres = 0.10,
  celltype.ident = "order"
)
```

@@

```{r Generate constrasts major groups for equine PBMC analysis}
# set major group/compartment vectors
B.clusters <- c(9, 15, 0, 22, 25, 27, 30)
basophils.clusters <- c(23)
cd3prf1neg.clusters <- c(2, 7, 8, 31, 16, 1, 13, 10, 21, 3, 6)
cd3prf1pos.clusters <- c(4, 5, 14, 17, 20)
dc.clusters <- c(19, 24, 26)
mono.clusters <- c(18, 11, 12, 28)
myeloid.clusters <- c(18, 11, 12, 28, 19, 24, 26, 29)

assignment.list <- list(
  B.clusters,
  basophils.clusters,
  cd3prf1neg.clusters,
  cd3prf1pos.clusters,
  dc.clusters,
  mono.clusters,
  myeloid.clusters
)

names(assignment.list) <- grep(ls(), pattern = "\\.clusters", value = T)

major.group.contrasts <-
  lapply(
    seq_along(assignment.list),
    function(x) {
      contrastConstructor(
        clusterIDs = assignment.list[[x]],
        design = edgeRobj$design,
        name = names(assignment.list)[[x]],
        makeContrast = TRUE
      )
    }
  )

major.group.contrasts <- do.call(cbind, major.group.contrasts)
colnames(major.group.contrasts) <- names(assignment.list)
```

```{r Generate constrasts for subsets within major groups for equine PBMC analysis}
assignment.list$basophils.clusters <- NULL

lapply(
  seq_along(assignment.list),
  function(x) {
    for (i in 1:length(assignment.list[[x]])) {
      contrastConstructor(
        clusterIDs = assignment.list[[x]][[i]],
        clusterIDs.2 = setdiff(assignment.list[[x]], assignment.list[[x]][[i]]),
        design = edgeRobj$design,
        makeContrast = T,
        name = paste(names(assignment.list)[[x]], "subset", assignment.list[[x]][[i]], sep = "_")
      )
    }
  }
)

subset.cluster.contrasts <- do.call(cbind, lapply(grep(ls(), pattern = "_subset_", value = T), get))
subset.cluster.contrasts <- as.data.frame(subset.cluster.contrasts)
colnames(subset.cluster.contrasts) <- grep(ls(), pattern = "_subset_", value = T)
```

@@

```{r Conduct differential gene expression testing for major cell groups and subsets}
major_dge_list <- list()
for (i in 1:length(colnames(major.group.contrasts))) {
  major_dge_list[[i]] <- glmQLFTest(edgeRobj$fit, contrast = major.group.contrasts[, i]) %>%
    topTags(n = NULL)
}

subset_dge_list <- list()
for (i in 1:length(colnames(subset.cluster.contrasts))) {
  subset_dge_list[[i]] <- glmQLFTest(edgeRobj$fit, contrast = subset.cluster.contrasts[, i]) %>%
    topTags(n = NULL)
}
```

## Part 6.4: Append percent.expression feature data and putative surface protein classification to differential gene expression tables

```{r Append percent.expression feature data and putative surface protein classification to differential gene expression tables}
major_dge_list <- add.percent.expression(
  dge_list = major_dge_list, object = ePBMC.integrated.obj,
  ident = "MajorGroup"
)

major.groups <- unique(ePBMC.integrated.obj$MajorGroup)

b.dge.list <- add.percent.expression(
  dge_list = mono.dge.list,
  object = ePBMC.integrated.obj,
  ident = "order",
  group.ident = "MajorGroup",
  subset = "B"
)

monodc.dge.list <- add.percent.expression(
  dge_list = mono.dge.list,
  object = ePBMC.integrated.obj,
  ident = "order",
  group.ident = "MajorGroup",
  subset = "Monocyte&DC"
)

basophil.dge.list <- add.percent.expression(
  dge_list = mono.dge.list,
  object = ePBMC.integrated.obj,
  ident = "order",
  group.ident = "MajorGroup",
  subset = "basophil"
)
```

@@

```{r Heatmap of DEGs between major cell groups}
major_tables <- major.glmQLFTest
major_tables <- major_tables[c("CD3PRF1neg", "CD3PRF1pos", "B", "Monocyte&DC", "Basophil")]
major_tables <-
  lapply(major_tables, tibble::remove_rownames) %>% lapply(tibble::column_to_rownames, "genes")
major.genes <- lapply(
  major_tables,
  function(df) {
    df[order(df$logFC, decreasing = T), ]
  }
) %>%
  lapply(rownames) %>%
  unlist2() %>%
  unique()

@@

```{r Stacked bars}
## Remove contaminating nuetrophils (low UMI)
Mcells.sub <- subset(Mcells, idents = 29, invert = T)
Mcells.sub$order <- droplevels(Mcells.sub$order)
horse_mono.dc.cluster_freq.table <- table(Mcells.sub$Subject, Mcells.sub$order) %>% melt()
colnames(horse_mono.dc.cluster_freq.table) <- c("Subject", "Celltype", "Fraction_of_all_cells")
horse_mono.dc.cluster_freq.table$Celltype <- as.factor(horse_mono.dc.cluster_freq.table$Celltype)

@@

```{r  Cross-species dendrogram between human and equine cell type clusters identified by scRNAseq}
# Set identity class to high-resolution clustering results
# Human
human_pbmc <- SetIdent(human_pbmc, value = "integrated_snn_res.1.2")
# Horse
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")

dendo <- cross_species_dendrogram(species.1.obj = ePBMC.integrated.obj, 
                                  species.2.obj = human_pbmc,
                                  species1.ident =  Mcells.order[-length(Mcells.order)], 
                                  species2.ident = c(14, 21, 15, 0, 1, 8, 22, 26, 19),
                                  species1.res = "order", 
                                  species2.res = "human.celltype",
                                  species1.name = "horse", 
                                  species2.name = "human")

@@

```{r Cross-species dendrogram between human and equine B cell type clusters identified by scRNAseq}
# Human
human_pbmc <- SetIdent(human_pbmc, value = "integrated_snn_res.1.2")
# Horse
ePBMC.integrated.obj <- SetIdent(ePBMC.integrated.obj, value = "order")

dendo <- cross_species_dendrogram(species.1.obj = ePBMC.integrated.obj, 
                                  species.2.obj = human_pbmc,
                                  species1.ident =  c(9, 15, 0, 22, 27, 30), 
                                  species2.ident = c(3, 11, 23),
                                  species1.res = "order", 
                                  species2.res = "human.celltype",
                                  species1.name = "horse", 
                                  species2.name = "human")
